%!TEX root = ../../dissertation.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Evaluating Reliable Streaming in Mobile Networks}
\label{chap:mobilestreaming-measurements}

Testing and evaluation plays a crucial role not just in correctly utilizing cross-layer data but also for media streaming in mobile networks in general. All of the discussed layer influences, interactions and other unintentional side-effects need to be thoroughly understood for the intended goal of optimizing reliable streaming in mobile networks.

The core network trace evaluations undertaken in Chapters~\ref{chap:mobilenets} and \ref{chap:mobilenetsmeasuring} were already able to highlight the load influencing dynamics of the core network control plane. But many more aspects remain uninvestigated, some of which may not even be covered by this kind of passive network-wide measurement type.

To further the understanding of the behavior of reliable streaming in mobile networks and investigate influence sources, usually either active measurements or simulations are conducted. Both methods and their drawbacks and benefits are discussed in this chapter. Moreover, an approach to enhance active measurements for mobile devices that facilitates meta-data of the device and improve measurement accuracy, is presented in the first section. Afterwards, a full \gls{LTE}-based reliable streaming simulation is also introduced and some initial results are given.



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\input{testbeds.tex}
\input{ltesim.tex}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\section{Summary}

It is safe to assume that reliable video streaming in mobile networks can be evaluated in a large number of ways. Besides the previously taken road of passively recorded network-scale traces, two other approaches are most valuable in investigating reliable streaming, active measurements and simulations. Both bring along their own set of benefits and drawbacks. 

Active measurements are better suited to test the wide range of influence factors present in actual networks and device protocol stacks. These can usually not be properly represented in a simulation. In order to increase the viability for mobile measurements, Sensorium was introduced, making it easier to record the precise state the mobile device currently is experiencing by giving access to all kinds of sensors.

If the goal is rather to investigate the theoretical overarching concepts of certain playback strategies or to scale up the test environment, one can utilize the simulative approach in the form of the \gls{LTE} mobile streaming simulation framework. Initial measurements using this framework already suggest a high influence of the network \gls{QoS} on the experienced streaming quality. A higher stalling probability caused by handovers during mobility was also uncovered.


%\section{Mobile Streaming and Measurement Summary}
%Cover both this and the previous Chapter?
%\todo{chapter summary required or fold it into thesis summary?}



%\url{http://www.wisebed.eu/}
%M-Lab \url{http://www.measurementlab.net/}

%Lessons Learned from the NetSense Smartphone Study 
%provided smartphones, subsidized by sprint
%just specific study, not a testbed

%OsmocomBB project \cite{osmocombbwww}


%An evaluation of dynamic adaptive streaming over \gls{HTTP} in vehicular environments \cite{Muller:2012:EDA:2151677.2151686}