%!TEX root = ../../dissertation.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Reliable Video Streaming in Mobile Networks}
\label{chap:mobilestreaming}

The analyses of the previous chapter were entirely based on having access to mobile core network measurements. Most research groups will in most likelihood not have this access and have to rely on active end-to-end device measurements or other approaches.

Furthermore, the notion of the importance of reliable video streaming in mobile networks was also not picked up in the previous chapter. Both the load characteristics of mobile networks as well as reliable streaming models were independently investigated up until now. But, as discussed in the beginning of this thesis and suggested in several mobile Internet traffic analyses and forecasts (see for example \cite{cisco2014VNI}), video is already occupying a large portion from mobile traffic and might rise to an even higher ratio in the near future.

Now that the two independent foundations were laid out in the previous chapters, the next two chapters once again merge mobile networks and reliable streaming with two further lines of discussions. 

First, the existing higher layer protocols of a mobile device's reliable video network stack and the influences of lower layer mobile layers on them are analyzed. Contrary to the popular scientific belief, that the Internet is unable to change and renew itself (so called Internet ``ossification'', compare, e.g., \cite{feldmann2010ossification}), protocols in the \gls{TCP}/\gls{IP} stack change all the time. Changes are more likely to occur in the end-to-end stack behavior and not in the packet format, as they are more difficult to change. New application requirements and changing network architectures always trigger adaptations in the stack in-between. Therefore, new and upcoming protocols are presented with an additional focus on the rapid changes to the existing network stack that occurred in the past years.

This also ties in with the mobile network load influencing factors discussed in Section~\ref{c4:sec:loaddefinition}. Any behavior exhibited in the device's stick could also have positive or negative effects on the network's control plane and signaling procedures. The section concludes with a theoretic cross-layer information exchange model that has the potential to improve reliable streaming in scenarios with fluctuating mobile network connections, e.g., during mobility, by informing the streaming application about it and making it possible to react upon such events.

Second, in Chapter~\ref{chap:mobilestreaming-measurements} now that the stack has been discussed fully, several approaches to monitor and measure end-to-end mobile network traffic, and video in particular, are demonstrated and executed. No access to in-network measurement probes limits the level of detail one can deduce from influence sources inside the network. But it also opens up a wide range of other device-based information sources, some of which are crucial in mobile devices, especially regarding mobility. These methods are also not limited to particular mobile network architectures and can thus be used to rapidly compare new architecture evolutions or specific core network implementations to each other. Finally, a full video streaming mobile network simulation model is also presented with some initial results.


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\input{enhancements.tex}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\input{crosslayer.tex}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Mobile Streaming Summary}

To summarize, application layer protocols are sensitive to circumstances of the lower protocol layers. Reliable streaming is no exception to this rule. Due to the time constraints present in media streaming it might even be more sensitive to outside influences than some other protocol with more relaxed timing constraints. Those protocol layer influences are aplenty, especially in mobile networks, and are most often unintended side effects of some protocol feature or behavior.

But this Chapter showed, that the mobile and Internet protocol stack is still very much in ongoing development and many of these side effects are known and being worked on. This happens either through the replacement of old protocols with entirely new versions, that attempt avoid these influences, or changes to existing protocols, as can be most prominently seen in the steady changes to \gls{TCP}. However, as it might not be enough to improve the individual layers, a concept for cross-layer cooperation between protocols on different layers is also highly desirable. The framework provided here should serve as a blueprint for future reliable applications to correctly facilitate information from the mobile network, e.g., handover event information.


