%!TEX root = ../../dissertation.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Evaluating Mobile Signaling Traffic and Load}
\label{chap:mobilenetsmeasuring}

With the architectural and methodological overview concluded this chapter can now move on to the actual evaluation. To this end, the previously described dataset is explored for any control plane load related signs. The evaluation, including a statistical analysis, is given in Section~\ref{c4:sec:evaluations}. The measurement data backs up a number of assumptions on the behavior of different device and operating system types, but also reveals some remarkable differences in signaling characteristics.

The results of these measurements are then used to construct a queuing theoretic load model for a \gls{CN} in Section~\ref{c4:sec:modeling}. This model is then extended with  virtualization modifications to it and followed by a numerical simulation in Section~\ref{c4:sec:simulation} to confirm the viability of the models.



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\input{evaluations.tex}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\input{modeling.tex}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\input{simulation.tex}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Core Network Evaluation Summary}
\label{c4:sec:conclusion}


The investigation of a week-long measurement trace recorded in an operational core network revealed some interesting signaling characteristics especially regarding the interdependency between user plane and control plane. Additionally, \gls{gtp} tunnel properties were determined to be a worthwhile measure for control plane load at the \gls{GGSN}, one of the central nodes in a \gls{3G} core network.

The investigation showed that the control plane is easily influenced by several device-based --- as far as they can be distinguished in a core network trace --- and time-of-day related features. The overall diurnal tunnel signaling load closely resembles the progression of the user plane. Most of the control plane's procedures are still triggered, either directly or indirectly, by user devices, of which the offered load is much smaller during night time. The trace evaluation also confirms the dominating influence of smartphones compared to other devices types, even when looking at the control plane.

But this also means that sheer traffic volume is not a good measure to determine load, as the per-device traffic volume of a smartphone is rather low when compared to devices like pure \gls{3G} modems attached to a notebook. In this aspect, the findings also support the stories of signaling storms in mobile networks caused by applications regularly causing small amounts of network traffic. Each application interaction results in disproportionate amounts of signaling load being generated. Even worse, measures taken to improve the radio interface control plane such as Fast Dormancy could possibly have adverse effects to core signaling as they might increase the tunnel churn.

But the load investigation should not stop here. The presented approaches were just the ones that could be conducted with the available data. If one were to have access to a mobile network monitoring system or more detailed data records from such a system, it would open up many more angles in the investigation. For example, recording every individual signaling message with all \glspl{IE} would give hard numbers on the direct signaling overhead, as could measurement probes located inside the network nodes report on the CPU and memory load in order to determine the control plane's processing overhead. A closer investigation of control plane load in relation to mobility behavior should also prove very interesting, as this is one of the central motifs in every mobile network.

Learning from this historical data, queuing theoretic models were created that can describe the control plane load in such networks. These models can be easily used in network dimensioning and planning processes by means of, e.g., stationary analyses. The novel baseline control plane load model presented here is a $M(t)/G(t)/c/0$ non-stationary Erlang loss model. When used in conjunction with parameters derived from the measurement traces it can easily be used for network dimensioning. To improve scaling in the future a further \gls{GGSN} load model with features used in virtualization was also proposed.

Due to general solvability issues of non-stationary Erlang models the model is evaluated and validated using a queuing simulation in terms of their blocking and tunnel state probability as well as the general resource utilization. The virtual model provided the added benefit of being more flexible in its scaling properties and energy efficiency. This might even lead to new \gls{GGSN}-as-a-Service business models, removing the need to provide and operate large amounts of infrastructure for rare cases of peak load. 

All of these properties serve to show the complexity of current mobile network systems even without running media streaming on top of it. Streaming in itself, while not being a real-time duplex communication protocol, is relatively sensitive to timings and influences from lower network layers, which can make streaming over mobile networks rather problematic. The next chapters investigate these issues and methods to evaluate them more closely.

%In the future we would like to deepen our modeling efforts to provide more dimensioning options for a core network. Also, we want to further investigate the correlation of user traffic and signaling and take a look at the implications specific traffic types bring for the core network. 

%All these investigations and modeling efforts combined could lead to a more informed approach of network planning: Being more aware of the control plane provides the necessary tools to identify probable causes for control plane activity. 

%We would also like to expand our evaluations, as there are several angles not investigated so far that could prove worthwhile. This includes an examination of the exact number and size of signaling messages flowing through the core, a more detailed picture of the processing load these messages induce at the \gls{GGSN}, and an evolved model. Furthermore, a differential analysis of our data compared to a newer dataset (potentially including \gls{LTE} access) could really prove worthwhile.

%We also look forward to searching for multiple active tunnels per device. As discussed previously, the \textit{Secondary PDP Context Activation Procedure} enables devices to establish up to ten additional tunnels attributed with a different, higher QoS level, if the network supports this. The additional load of managing and holding multiple tunnels plus the displacement of other, ``lower-quality'' traffic could prove to be an interesting investigation. Initial observations indicate that this feature is rarely used today by very few types of devices, but it will be of increased interest in the face of ongoing LTE/EPS deployments, whose specifications expand upon this secondary tunnel concept.